Testing for measurement invariance with respect to an ordinal variable

### 2013-04-16 ordinal/response.txt

Comment 2.1. On page 5 is the scaled sum of scores which is normalized by the square root of n and the covariance matrix of the scores I. This is fine theoretically, but what choice do you make in practice? In fact I think that here the problem is model misspecification, other than the biasing factor V. If the model used in ML is incorrect with respect to the number of factors for instance, then this will result in a inconsistent estimate of I. Why not use the sandwich by the late Halbert White (JASA, 1982 for instance)? This is easy to calculate and will improve the limiting distribution of the test statistics.  The choice of I is also missing in the simulations. Please be more specific what you did there.

Response 2.1. We have added an extended discussion of choices for the covariance matrix of scores, with references.  We have also clarified that, in the simulations, we used the information matrix.  Because our simulation models were correctly specified (except for the measurement invariance violation), we do not expect the choice of I to play a large role there.

Comment 2.2. The simulations show great improvement over the standard LR test on measurement invariance. And on page 10 the authors mention the results by Bentler and Bonett (1980) on LR tests picking up deviations too quickly. I think the alternative tests, adjustments of LR tests, like the one proposed in Yuan and Bentler (JASA, 1997). There the LR is downweighted ate rate O_p(1/n), which brings the results of the LR closer to the results of the authors' maxLM test. I would like to see a comparison with this or maybe some other adjusted LR test.

Response 2.2. We implemented three adjusted LRTs in our simulations: (i) the Yuan-Bentler 1997 adjustment suggested by the reviewer, (ii) the use of AIC (which, for nested models, essentially increases the critical value of the LRT), and (iii) the Satorra-Bentler adjustment with correction for difference tests.  While these adjustments downweight the LRT associated with individual models (e.g., comparing an individual model to a saturated model), the impact of the adjustments on difference tests is less clear.  Further, the Yuan-Bentler adjustment is seemingly designed for situations where the model is estimated via WLS, and there are no available corrections for difference tests.  We suspect that these issues are problematic, as our simulation results indicated that these statistics are worse than the usual LRT in nearly all conditions.  We have included graphs of the results with the resubmission but have not included these results in the manuscript.  We are willing to reconsider the exclusion of these results if the AE or reviewers find them of interest.

### 2013-04-16 ordinal/ordinal.Rnw
Revisions to ordinal ms: o Added detail to re-ordered introductory sections o Small revisions to response letter
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@197 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

In studying measurement invariance, we are essentially testing the
assumption that all  
individuals arise from the same 
parameter vector ${\bm \theta}$.
Thus, a hypothesis of invariance can be written as
\begin{equation}
    \label{eq:h0}
    H_0:~ {\bm \theta}_i = {\bm \theta}_0,\quad (i=1,\ldots,n),
\end{equation}
where ${\bm \theta}_i$ reflects the parameter vector for individual
$i$ (and modifications for subsets of ${\bm \theta}$ are immediate).  
The most general alternative hypothesis related to $V$ may then be
written as
\begin{equation}
  \label{eq:h2*}
  H_1^*:~ {\bm \theta}_i = {\bm \theta}_{v_i},
\end{equation}
stating that the parameter vector differs for every unique realization
of $V$.  This alternative is commonly employed when $V$ is
categorical.  In these situations, the likelihood ratio test (LRT) compares
a model where 
parameters are restricted across groups (i.e., across values of $V$)
to a model where parameters are free across groups; the exact
parameter values within each group are completely unrestricted.
However, in situations where $V$ is categorical or
continuous,~\eqref{eq:h2*} includes non-monotonic violations of
measurement invariance.   This allows for instances where, e.g., the
parameter values initially increase with $V$ and then decrease, 
or where just one or two ``middle'' levels of $V$ differ from the rest.
Researchers typically do not expect such a result when testing
measurement invariance w.r.t.\ continuous or ordinal $V$, and 
researchers often cannot interpret such violations.  Monotonic
parameter changes w.r.t.\ $V$ are of much more interest in these
situations, with the simplest type of change given by the alternative
hypothesis
\begin{equation}
  \label{eq:h1*}
  H_2^*:~ {\bm \theta}_i = \left\{ \begin{array}{ll}
    {\bm \theta}^{(A)} & \mbox{if } v_i \le \nu, \\
    {\bm \theta}^{(B)} & \mbox{if } v_i >   \nu,
  \end{array} \right.
\end{equation}
where $\nu$ is a threshold dividing individuals into two groups based on
$V$.  This alternative is implicitly employed in ``median split''
analyses, 
where $\nu$ is given as the sample median of $V$.  The threshold $\nu$
is usually unknown, however, so it is generally of interest to
test~\eqref{eq:h1*} across all possible values of $\nu$.  The tests
proposed below generally allow for this.

@@

We  consider situations in which a $p$-dimensional variable $X$
with observations $\bm{x}_i, i=1,\dots,n$ is described by a 
model with density $f(\bm{x}_i; \bm{\theta})$ and
associated joint log-likelihood
\begin{equation} \label{eq:loglik}
  \ell(\bm{\theta}; \bm{x}_1, \dots, \bm{x}_n) ~=~
    \sum_{i = 1}^n \ell(\bm{\theta}; \bm{x}_i) ~=~
    \sum_{i = 1}^n  \log f(\bm{x}_i; \bm{\theta}),
\end{equation}
where ${\bm \theta}$ is some $k$-dimensional parameter vector that characterizes
the distribution.

We focus on applications where the density $f(\bm{x}_i; \bm{\theta})$ arises
from a structural equation model with assumed multivariate normality,
though the proposed tests extend beyond this family of models.  Under
the usual regularity conditions (e.g., \citeNP{Fer96}), the model parameters
$\bm{\theta}$ can 
be estimated by maximum likelihood (ML), i.e.,
\begin{equation} \label{eq:ml}
  \hat{\bm{\theta}} ~=~ \argmax_{\bm{\theta}} \ell(\bm{\theta}; x_1, \dots, x_n),
\end{equation}
or equivalently by solving the first order conditions
\begin{equation}
    \label{eq:ml1}
  \sum_{i=1}^{n} {\bm s}(\hat{\bm{\theta}}; \bm{x}_i) ~=~ 0,    
\end{equation}
where 
\begin{equation}
  \label{eq:score}
  {\bm s}({\bm \theta}; x_i) ~=~ \left(
    \frac{\partial \ell({\bm \theta}; x_i)}{\partial \theta_1},
    \dots,
    \frac{\partial \ell({\bm \theta}; x_i)}{\partial \theta_k}
  \right)^\top,
\end{equation}
is the score function of the model (the partial derivative of the casewise
likelihood contributions w.r.t.\ the parameters $\bm{\theta}$).
Evaluation of the score function at $\hat{\bm{\theta}}$ for $i=1,
\dots, n$ essentially measures the extent to which the model maximizes
each individual's likelihood: as an individual's scores stray further
from zero, the model provides a poorer description of that individual.

@@

The four ``traditional'' statistics that we consider generally treat
the ordinal auxiliary variable as categorical.  The first two are the
likelihood ratio test of measurement 
invariance in the six unique variance parameters and the
unordered LM test from~\eqref{eq:lmuo}.  Next, we consider the
Satorra-Bentler \citeyear{SatBen01} scaled likelihood ratio test
statistic with correction for difference testing.  Finally, we
consider the use of AIC \cite{Aka74} for detecting measurement
invariance.  In this context, we compute AIC for both the ``reduced'' 
model (all parameters constrained equal across groups) and the
``full'' model (unique variances allowed to differ between groups).
We then declare a measurement invariance violation if the AIC for the
full model is lower than the AIC for the reduced model.

@@

Simulation results comparing the ordinal tests to the Satorra-Bentler
scaled test statistic and to the AIC are presented in
Figure~\ref{fig:sim1bres}.  This figure has the same design as the
previous figure.  It is again seen that the $\max \mathit{LM}_o$
statistic exhibits the greatest power across the conditions.
Additionally, the Satorra-Bentler statistic has very large Type I
error rates at low-$n$/large-$m$ combinations but is comparable to the
ordinal double-max test at large $n$.  Finally, the ``power'' of the
AIC (measured by the proportion of times that the full model has a
lower AIC than the reduced model) is exceptionally low across all
conditions.  The full model's gain in fit is seldom worth the extra
parameters, at least according to AIC.

@@

\begin{figure}
\caption{Simulated power curves for the ordered 
      $\max\mathit{LM}$ tests, the ordered double-max test, the
      Satorra-Bentler (2001) scaled likelihood ratio
      test, and the use of AIC 
      across three sample sizes $n$, three 
      levels of the ordinal variable $m$, and measurement invariance
      violations of 0--1.5 standard errors (scaled by $\sqrt{n}$).}
\label{fig:sim1bres}
\setkeys{Gin}{width=\textwidth}
<<sim1bres, fig=TRUE, height=7, width=7>>=
sim1.tmp <- sim1[sim1$test %in% c("maxLM_o","WDM_o","LRT_sb","AIC"),]
sim1.tmp$test <- factor(sim1.tmp$test)
trellis.par.set(theme = canonical.theme(color = FALSE))
mykey <- simpleKey(levels(sim1.tmp$test), points = TRUE, lines = TRUE)
print(xyplot(power ~ diff | nlevels + nobs, group = ~ test, data = sim1.tmp, 
             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey))
@
\end{figure}

@@

Simulation results for the two ordinal test statistics, the Satorra-Bentler
scaled test statistic, and the AIC are presented in
Figure~\ref{fig:sim2bres}.  It is observed that the Satorra-Bentler
statistic exhibits performance that is similar to the usual LRT,
with the two ordinal statistics having lower power.  The AIC, which
generally exhibited low power even for Simulation 1, continues to
exhibit low power here.  However, if $n$ or $d$ is large enough, the
power exhibited by AIC is larger than the power of the ordinal test
statistics.

@@

\begin{figure}
\caption{Simulated power curves for the ordered 
      $\max\mathit{LM}$ tests, the ordered double-max test, the
      Satorra-Bentler (2001) scaled likelihood ratio
      test, and the use of AIC 
      across three sample sizes $n$, three
      levels of the ordinal variable $m$, and measurement invariance
      violations of 0--3 standard errors (scaled by $\sqrt{n}$)
      occurring at a single level (the $(1 +
      m/2)^{\text{th}}$ level) of the ordinal auxiliary variable.}
\label{fig:sim2bres}
\setkeys{Gin}{width=\textwidth}
<<sim2bres, fig=TRUE, height=7, width=7>>=
sim2.tmp <- sim2[sim2$test %in% c("maxLM_o","WDM_o","LRT_sb","AIC"),]
sim2.tmp$test <- factor(sim2.tmp$test)
trellis.par.set(theme = canonical.theme(color = FALSE))
mykey <- simpleKey(levels(sim2.tmp$test), points = TRUE, lines = TRUE)
print(xyplot(power ~ diff | nlevels + nobs, group = ~ test, data = sim2.tmp, 
             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey))
@
\end{figure}

>>>>

In studying measurement invariance, we consider situations in which a $p$-dimensional variable $X$
with observations $\bm{x}_i, i=1,\dots,n$ is described by a 
model with density $f(\bm{x}_i; \bm{\theta})$ and
associated joint log-likelihood
\begin{equation} \label{eq:loglik}
  \ell(\bm{\theta}; \bm{x}_1, \dots, \bm{x}_n) ~=~
    \sum_{i = 1}^n \ell(\bm{\theta}; \bm{x}_i) ~=~
    \sum_{i = 1}^n  \log f(\bm{x}_i; \bm{\theta}),
\end{equation}
where ${\bm \theta}$ is some $k$-dimensional parameter vector that characterizes
the distribution.

Tests of  measurement invariance are essentially tests of the 
assumption that all  
individuals arise from the same 
parameter vector ${\bm \theta}$.
Thus, a hypothesis of measurement invariance can be written as
\begin{equation}
    \label{eq:h0}
    H_0:~ {\bm \theta}_i = {\bm \theta}_0,\quad (i=1,\ldots,n),
\end{equation}
where ${\bm \theta}_i$ reflects the parameter vector for individual
$i$ (and modifications for subsets of ${\bm \theta}$ are immediate).  
The most general alternative hypothesis related to $V$ may then be
written as
\begin{equation}
  \label{eq:h2*}
  H_1^*:~ {\bm \theta}_i = {\bm \theta}_{v_i},
\end{equation}
stating that the parameter vector differs for every unique realization
of $V$.  This alternative is commonly employed when $V$ is
categorical.  In these situations, the likelihood ratio test (LRT) compares
a model where 
parameters are restricted across groups (i.e., across values of $V$)
to a model where parameters are free across groups; the exact
parameter values within each group are completely unrestricted.
However, in situations where $V$ is categorical or
continuous,~\eqref{eq:h2*} includes non-monotonic violations of
measurement invariance.   This allows for instances where, e.g., the
parameter values initially increase with $V$ and then decrease, 
or where just one or two ``middle'' levels of $V$ differ from the rest.
Researchers typically do not expect such a result when testing
measurement invariance w.r.t.\ continuous or ordinal $V$, and 
researchers often cannot interpret such violations.  Monotonic
parameter changes w.r.t.\ $V$ are of much more interest in these
situations, with the simplest type of change given by the alternative
hypothesis
\begin{equation}
  \label{eq:h1*}
  H_2^*:~ {\bm \theta}_i = \left\{ \begin{array}{ll}
    {\bm \theta}^{(A)} & \mbox{if } v_i \le \nu, \\
    {\bm \theta}^{(B)} & \mbox{if } v_i >   \nu,
  \end{array} \right.
\end{equation}
where $\nu$ is a threshold dividing individuals into two groups based on
$V$.  This alternative is implicitly employed in ``median split''
analyses, 
where $\nu$ is given as the sample median of $V$.  The threshold $\nu$
is usually unknown, however, so it is generally of interest to
test~\eqref{eq:h1*} across all possible values of $\nu$.  The tests
proposed below generally allow for this.

@@

\subsection{Model Estimation}
We focus specifically on applications where the density $f(\bm{x}_i; \bm{\theta})$ arises 
from a structural equation model with assumed multivariate normality,
though the proposed tests extend beyond this family of models.
  Under
the usual regularity conditions (e.g., \citeNP{Fer96}), the model parameters
$\bm{\theta}$ can 
be estimated by maximum likelihood (ML), i.e.,
\begin{equation} \label{eq:ml}
  \hat{\bm{\theta}} ~=~ \argmax_{\bm{\theta}} \ell(\bm{\theta}; x_1, \dots, x_n),
\end{equation}
or equivalently by solving the first order conditions
\begin{equation}
    \label{eq:ml1}
  \sum_{i=1}^{n} {\bm s}(\hat{\bm{\theta}}; \bm{x}_i) ~=~ 0,    
\end{equation}
where 
\begin{equation}
  \label{eq:score}
  {\bm s}({\bm \theta}; x_i) ~=~ \left(
    \frac{\partial \ell({\bm \theta}; x_i)}{\partial \theta_1},
    \dots,
    \frac{\partial \ell({\bm \theta}; x_i)}{\partial \theta_k}
  \right)^\top,
\end{equation}
is the score function of the model (the partial derivative of the casewise
likelihood contributions w.r.t.\ the parameters $\bm{\theta}$).
Evaluation of the score function at $\hat{\bm{\theta}}$ for $i=1,
\dots, n$ measures the extent to which the model maximizes
each individual's likelihood: as an individual's scores stray further
from zero, the model provides a poorer description of that individual.

@@

The two ``traditional'' statistics that we consider generally treat
the ordinal auxiliary variable as categorical.  These include the
likelihood ratio test of measurement 
invariance in the six unique variance parameters and the
unordered LM test from~\eqref{eq:lmuo}.  At the request of reviewers,
we also considered the
Satorra-Bentler \citeyear{SatBen01} scaled likelihood ratio test
statistic with correction for difference testing, the Yuan-Bentler
\citeyear{YuaBen97} scaled test statistic, and the use of AIC
\cite{Aka74} for detecting measurement 
invariance.  We do not report the latter results, because these test
statistics performed worse than the usual likelihood ratio test (for
both Simulation 1 and Simulation 2).

%% In this context, we compute AIC for both the ``reduced'' 
%% model (all parameters constrained equal across groups) and the
%% ``full'' model (unique variances allowed to differ between groups).
%% We then declare a measurement invariance violation if the AIC for the
%% full model is lower than the AIC for the reduced model.

@@

%% Simulation results comparing the ordinal tests to the Satorra-Bentler
%% scaled test statistic and to the AIC are presented in
%% Figure~\ref{fig:sim1bres}.  This figure has the same design as the
%% previous figure.  It is again seen that the $\max \mathit{LM}_o$
%% statistic exhibits the greatest power across the conditions.
%% Additionally, the Satorra-Bentler statistic has very large Type I
%% error rates at low-$n$/large-$m$ combinations but is comparable to the
%% ordinal double-max test at large $n$.  Finally, the ``power'' of the
%% AIC (measured by the proportion of times that the full model has a
%% lower AIC than the reduced model) is exceptionally low across all
%% conditions.  The full model's gain in fit is seldom worth the extra
%% parameters, at least according to AIC.

@@

%% \begin{figure}
%% \caption{Simulated power curves for the ordered 
%%       $\max\mathit{LM}$ tests, the ordered double-max test, the
%%       Satorra-Bentler (2001) scaled likelihood ratio
%%       test, and the use of AIC 
%%       across three sample sizes $n$, three 
%%       levels of the ordinal variable $m$, and measurement invariance
%%       violations of 0--1.5 standard errors (scaled by $\sqrt{n}$).}
%% \label{fig:sim1bres}
%%\setkeys{Gin}{width=\textwidth}
%%<<sim1bres, fig=TRUE, height=7, width=7>>=
%%sim1.tmp <- sim1[sim1$test %in% c("maxLM_o","WDM_o","LRT_sb","AIC"),]
%%sim1.tmp$test <- factor(sim1.tmp$test)
%%trellis.par.set(theme = canonical.theme(color = FALSE))
%%mykey <- simpleKey(levels(sim1.tmp$test), points = TRUE, lines = TRUE)
%%print(xyplot(power ~ diff | nlevels + nobs, group = ~ test, data = sim1.tmp, 
%%             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey))
%%@
%%\end{figure}

@@

%% Simulation results for the two ordinal test statistics, the Satorra-Bentler
%% scaled test statistic, and the AIC are presented in
%% Figure~\ref{fig:sim2bres}.  It is observed that the Satorra-Bentler
%% statistic exhibits performance that is similar to the usual LRT,
%% with the two ordinal statistics having lower power.  The AIC, which
%% generally exhibited low power even for Simulation 1, continues to
%% exhibit low power here.  However, if $n$ or $d$ is large enough, the
%% power exhibited by AIC is larger than the power of the ordinal test
%% statistics.

@@

%% \begin{figure}
%% \caption{Simulated power curves for the ordered 
%%       $\max\mathit{LM}$ tests, the ordered double-max test, the
%%       Satorra-Bentler (2001) scaled likelihood ratio
%%       test, and the use of AIC 
%%       across three sample sizes $n$, three
%%       levels of the ordinal variable $m$, and measurement invariance
%%       violations of 0--3 standard errors (scaled by $\sqrt{n}$)
%%       occurring at a single level (the $(1 +
%%       m/2)^{\text{th}}$ level) of the ordinal auxiliary variable.}
%% \label{fig:sim2bres}
%%\setkeys{Gin}{width=\textwidth}
%%<<sim2bres, fig=TRUE, height=7, width=7>>=
%%sim2.tmp <- sim2[sim2$test %in% c("maxLM_o","WDM_o","LRT_sb","AIC"),]
%%sim2.tmp$test <- factor(sim2.tmp$test)
%%trellis.par.set(theme = canonical.theme(color = FALSE))
%%mykey <- simpleKey(levels(sim2.tmp$test), points = TRUE, lines = TRUE)
%%print(xyplot(power ~ diff | nlevels + nobs, group = ~ test, data = sim2.tmp, 
%%             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey))
%%@
%%\end{figure}
